---
title: "ALB Spread Simulation"
output: pdf_document
author: "Lora Murphy"
linkcolor: blue
urlcolor: blue
editor_options: 
  markdown: 
    wrap: 72
---

The ALB spread simulation takes a set of trees with an initial ALB
infestation, and models insect spread for a desired length of time.

# Required setup

The APHIS simulation is performed in R, and takes advantage of R's
spatial analysis capabilities. (More on R and spatial data at
<https://cran.r-project.org/view=Spatial>.

To get set up:

1.  Install the latest version of R. <https://cran.r-project.org/>

2.  If you are working on a computer running Windows, install RTools,
    which can be found under the R Binaries section of the website
    above. Follow the directions to make sure R and RTools can find each
    other.

3.  Install the additional packages "Rcpp", "sp" and "sf". Be sure to
    use dependencies=TRUE when you do:

```{r eval=FALSE}
install.packages(c("sp", "sf", "Rcpp"), dependencies=TRUE)
```

4.  Additionally, the packages "magick", "ggplot2", and "ggpubr" may be
    required if you want to use functions in the provided graphics
    script (not required to perform simulations).

# How to run a simulation

Open the script called "Simulation.R". Adjust the settings you see there
as desired, and run in R.

Best practice is to save a copy of this script for each run you do. This
provides a record of your settings.

The other scripts do the heavy lifting. They will take their settings
from "Simulation.R" and you shouldn't need to adjust them to run the
model.

# Input

The input is a dataframe of trees that are potential ALB victims. All
are assumed to belong to the genus *Acer*. The dataframe should have the
following fields, with these exact case-sensitive names (there can be
other fields, they'll be ignored):

-   "x", "y": location coordinates. Point of origin doesn't matter, this
    will only be used to calculate distances between points. SPECIFY THE
    LINEAR UNITS BELOW (feet, meters).
-   "mean_noforestdist": the mean distance to the forest edges in each
    of the 8 cardinal and intercardinal directions, in m. If the tree
    isn't in a forest landcover, this is 0.
-   "dbh": tree's DBH in cm
-   OPTIONAL: "infested": integer, 0 being uninfested, 1 being infested.
    If this field is included and there is at least 1 tree with a 1
    status, the trees marked as 1 will be the seed of the outbreak. If
    either this field is not included or all the values are 0, a select
    number of trees will be randomly assigned to be the seed of the
    outbreak (number settable below).
-   OPTIONAL: "unit": identifier of any format, that can be used to
    prescribe surveys.

The "mean_noforestdist" field can be calculated with the script "Get non
forest landcover in 8 directions.R."

OPTIONAL INPUT: A dataframe that specifies what surveys to do, and when
to do them. Two required fields: "year", with the year to do the survey
(matching a year in the range you specify for starting and ending
years), and "unit", an identifier that matches the "unit" field in the
trees dataframe.

# Output

Two objects get saved for every run: the trees, and the surveys done.

## Trees

The trees are saved as a SpatialPointsDataFrame that is a copy of the
input dataframe, with additional fields for dates of infestation and
removal ("year_infested" and "year_removed").

A SpatialPointsDataFrame is a dataframe with spatial information for
each line, equivalent to a points shapefile in ArcGIS. In most cases,
dataframe syntax can be used on it directly; for instance, the \$
operator will work as expected. It can be coerced into a regular
dataframe using as.data.frame.

The SpatialPointsDataFrame can be easily be plotted as a set of points
on a map using plot (for example: plot(trees) is all that's necessary),
or used in other spatial analyses. It can also be written out as a
shapefile compatible with a variety of GIS software using the "rgdal"
package in R.

## Surveys

The surveys performed (if any) are saved as a list, with one slot for
each year of the simulation. An empty list slot indicates that no survey
occurred that year. If a survey did occur, it is saved as a
SpatialPolygons object with one or more polygons indicating the surveyed
regions. You can visualize the surveys using the plot command (see
included graphics file for more detailed examples).

The trees and surveys are each saved in their own RDS file by default.
This stores them in a native R format. Read them back in using the
readRDS command, assigning whatever name you like. For example:

```{r eval = FALSE}
trees <- readRDS("mytrees.rds")
surveys <- readRDS("mysurveys.rds")
plot(trees)
plot(surveys[[3]], add=T)
```

# Management

Infested trees are (mostly) detected during surveys. Surveys can be
managed in one of two ways: the simulation can automatically decide what
is surveyed based on detections, or you can explicitly specify areas to
be surveyed and when.

If the simulation is automatically determining surveys, a survey will
occur in a circle centered on each infested tree detected in the
previous year, of a radius to be set by the user. Circles may overlap.

If you want to specify when and where surveys will be done, set up a
dataframe with two fields: "year" and "unit". The trees dataframe should
also have a field called "unit". For example:

| year | unit  |
|------|-------|
| 1    | ZN_1  |
| 2    | ZN_5  |
| 2    | ZN_6  |
| 4    | ZN_10 |

In year 2, all trees whose unit equals "ZN_5" or "ZN_6" will be
surveyed. No survey will occur in year 3.

# Model flow

1.  Choose the risk model and do any necessary setup.

2.  Loop through the years of the simulation:

    a.  Calculate this year's probability of infestation for potential
        hosts, based on the chosen risk model.

    b.  Choose the trees that get infested this year. For each tree,
        compare a random number to its probability to make the choice.

    c.  For those trees that get infested, record the year they get
        infested to both track time until emergence, and generally track
        spread in output.

    d.  Apply management. As parameters, 4 different probabilities are
        defined for detection and removal: post-emergence tree in
        surveyed area, pre-emergence tree in surveyed area,
        post-emergence tree NOT in surveyed area, pre-emergence tree NOT
        in surveyed area.

    e.  Decide what's being surveyed: either decide automatically based
        on trees that were detected in the previous year, or check the
        list of surveys to be done and choose all trees in units to be
        surveyed in the current year. It is possible, with either
        method, that there will be no surveys in a given year.

    f.  Identify all post-emergence trees within surveyed areas, if any,
        and use a random number against the probability of detection of
        these trees to choose which are detected and removed.

    g.  If the removal budget has not been met, identify all
        pre-emergence infested trees within surveyed areas. Use a random
        number on each, compared to the detection probability, to choose
        which got detected and removed.

    h.  Repeat steps b and c, with non-surveyed areas and post- and
        pre-emergence trees. All detected trees will be removed, up to
        the budget.

    i.  Apply tree growth to all live trees.

3.  Clean up; create graphics; save output files.

# Risk model: determining which trees are infested

Uninfested trees have a certain probability of being infested each
timestep, as a function of number of and distance to post-emergence
infested source trees and various tree characteristics (see below).

Source trees are trees which are infested, post-emergence, and have not
been removed. A tree which is infested is pre-emergence until the lag
period has passed (see biology settings of the model). Pre-emergence
trees do not infest other trees. The lag period could be set to 0 to
remove the pre- and post-emergence distinction.

There are multiple possible risk models. This simulator is the result of
analyses conducted in several different geographic regions in the U.S.
These regions had different risk models that best captured their local
conditions and landscapes.

## Risk model parameters

Each risk model requires several parameter values to control its
behavior. The risk models were parameterized with local data in each
geographic region. Several parameter sets are included in the file
"Parameters.R". If you wish to use any of them, copy and paste them into
the main simulation script.

## Risk model 13

This model assesses tree risk as a function of source pressure, DBH of
tree, number of nearby *Acer* neighbors, and landcover. This model best
described conditions in Long Island, NY.

The probability that a tree will be infested is:
$$prob = \mu * source.term * dbh.term * density.term * distance.term$$
where $\mu$ is a parameter representing the upper limit of infestation
probability.

$$source.term = \cfrac{1}{1 + \left(\cfrac{sp}{ \alpha }\right)^\gamma} $$
where sp is a measure of source pressure and $\alpha$ and $\gamma$ are
parameters.

$$sp = \sum_{i=1}^N \beta * e^{(dsn*(distance_i ^ \delta))}$$ for N
source trees within the maximum distance of 5280 feet, where $dsn$,
$\beta$, and $\delta$ are parameters and $distance_i$ is the distance
from the source tree to the ith neighor.

$$dbh.term = \cfrac{1}{1 + \left(\cfrac{DBH}{ b_1 }\right)^{b_2}} $$
where DBH is tree's DBH, in cm, and $b_1$ and $b_2$ are parameters.

$$density.term = \cfrac{1}{1 + \left(\cfrac{N_{acer}}{ c_1 }\right)^{c_2}} $$
where $N_{acer}$ is the number of *Acer* neighbors within 30 m of the
target (will be calculated and updated by the script) and $c_1$ and
$c_2$ are parameters.

$$distance.term = \cfrac{1}{1 + \left(\cfrac{dist_{lc}}{ e_1 }\right)^{e_2}} $$
where $dist_lc$ is the mean distance to no-forest landcover and $e_1$
and $e_2$ are parameters.

## Risk model 11

This model assesses tree risk as a function of source pressure, DBH of
tree, number of nearby *Acer* neighbors, and landcover. This was the
best model in New York City and Worcester, MA.

The probability that a tree will be infested is:
$$prob = \mu * source.term * dbh.term * density.term * distance.term$$
where $\mu$ is a parameter representing the upper limit of infestation
probability.

$$source.term = \cfrac{1}{1 + \left(\cfrac{sp}{ \alpha }\right)^\gamma} $$
where sp is a measure of source pressure and $\alpha$ and $\gamma$ are
parameters.

$$sp = \sum_{i=1}^N \beta * dir_{i} * (d_1 + d_2)$$ for N source trees
within the maximum distance of 5280 feet, where $\beta$ is a parameter,
$dir_i$ is a parameter for the effect of direction from the tree to
source *i*, in increments of 30 degrees, north 0 and increasing in the
clockwise direction (0-1). $d_1$ and $d_2$ are separate near and far
distance kernels.

$$d_1 = \delta_1 * exp(\cfrac{dsn}{1000} * dist_i)$$
$$d_2 = \delta_2 * exp\left(-0.5 * \left(\cfrac{ln(\frac{dist_i}{dof})}{dbf}\right)^2\right)$$
where $\delta_1, \delta_2, dsn, dof$ and *dbf* are parameters, and
$dist_i$ is the distance to the source tree.

$$dbh.term = \cfrac{1}{1 + \left(\cfrac{DBH}{ b_1 }\right)^{b_2}}$$
where DBH is tree's DBH, in cm, and $b_1$ and $b_2$ are parameters.

$$density.term = \cfrac{1}{1 + \left(\cfrac{N_{acer}}{ c_1 }\right)^{c_2}}$$
where $N_{acer}$ is the number of *Acer* neighbors within 30 m of the
target (will be calculated and updated by the script) and $c_1$ and
$c_2$ are parameters.

$$distance.term = \cfrac{1}{1 + \left(\cfrac{dist_{lc}}{ e_1 }\right)^{e_2}}$$
where $dist_lc$ is the mean distance to no-forest landcover and $e_1$
and $e_2$ are parameters.

# Tree growth

In the case of a long simulation, growth may change a tree's risk over
time. This script can grow trees using a linear function of DBH:

$$DBH_{t+1} = DBH_t + (m*DBH_t + b)$$ where this year's DBH is
$DBH_{t+1}$, last year's DBH is $DBH_t$, $m$ is the slope of growth, and
$b$ is the intercept. Either $m$ or $b$ can be set to 0.
